On the eletrodynamic of moving bodies

### 2019-07-18 Article_AS.Rmd
Reorganisation intro

Thousands hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019]. It is important to increase tree growth to meet the growing demand of forest industry. The reforestation, when under human control, could ideally be made by using better trees than the former ones, i.e. trees that growth faster, that are more resistant to perturbation and that are of better structural quality. To achieve this task, we must take into account the cambial activity that is governed by environmental conditions and genetics [@Dufour2010; @Rossi2012]. Therefore, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@Bonfils2004]. In forestry, genetic selection is a process allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for human use [@zobel1984; @DRF2000; @nanson2004]. \

During the first half of the 20^th^ century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012]. Hence, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties. For example, improvement programs by genetic selection are in place for white spruce *(Picea glauca [Moench] Voss)* [@DRF2000; @Mullin2011] which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015] in North America. Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008]. In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@nienstaedt1990]. All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009]. In Quebec, several generations of seed orchards have already been established in white spruce. Thus, the current objectives of genetic selection for this species are to select trees with a higher wood quality  because white spruce has already reached a level considered interesting for traits such as rapid growth and volume of wood. \

@mitchell1961 defined the quality of the wood as follows: "Wood quality is the resultant of physical and chemical characteristics possessed by a tree or a part of a tree that enable it to meet the property requirements for different end products". it is composed mainly of two components: economical and biological (stem vigor). These two components are threfore two desirable traits that are taken into account in genetic selection. Biologically, resistance to mortality by vigor is important to ensure the survival of the tree. Economically, the value of the wood for end-use is important for industries. Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006]. More specifically the low interannual variability of wood density and rings width has an high importance for stem vigor and wood quality because these two traits are related to it. In fact, the density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a low variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008]. For example, @suarez2004 found that at the tree scale, individuals with variable growth were more prone to die of drought than more regularly growing trees. These studies have shown that the steady growth and therefore the low interannual variation wood density allows the tree to be more vigorous and therefore less prone to related mortality and environmental conditions. With regard to the economical context, interannual variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties. Indeed, low interannual variability of wood density affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002]. For example, the low interannual variation in wood density is appreciated for sawnwood because sawyers place a high value on the low interannual variation wood density [Hassegawa et al (pending approval); @X]. It is therefore important from both a biological and an end-use points of view to have a wood with a low interannual density variation within the stem. We understand that there might be a strong gain at selecting trees that are prone to produce wood with a low interannual variability of wood density to get a resistant and highly valuable forest in the next decades. For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Beaulieu2009; @Mullin2011]. For wood quality, one of the characteristics that one wishes to select is the amplitude of the interannual variation of wood density, provided that it has sufficiently interesting heritability for a genetic selection. \

Heritability is the probability that a trait of an individual is hereditarily transmitted by genetic factors. It is a statistical datum that serves to evaluate the part of genetic factors and that of environmental factors in the construction of the phenotype of a given population. Several studies have been set up to understand the genetic sources of phenotypic variations of wood quality traits and their heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016]. Thus, previous studies have shown that wood density is a trait under strong genetic control therefore strong heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006]. It has been proven that the wood density is under genetic control, but we can ask ourselves the question of its amplitude of interannual variation. The interannual variation of wood density comes from two sources: the low-frequency radial profile and the high-frequency radial profile [@Nicault2010]. The low-frequency radial profile is the radial variation of the density of pith-to-bark with a  high density during the first years of tree growth  decreases to a certain level and begins to increase again. In white spruce, this radial profile is the radial type II profile described by [@panshin1980]. These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] because this functioning is internal to the tree and is transmitted to future generations. The high-frequency radial profile is the amplitude of the interannual variation of wood density. We did not find any studies on heritability of amplitude of interannual variation of wood density in white spruce.\

Studies have linked annual environmental conditions such as soil water availability, climatic conditions that occur at a high frequency, and interannual variability wood density [@becker1995; @huang2007; @Nicault2010; @zaehle2014]. Indeed, long-term growth analyzes have shown a link between wood density and local annual conditions such as summer temperature in white spruce [@darrigo1992] and other species [@becker1995; @huang2007; @zaehle2014]. However, it seems to have some genetic control in the density sensitivity to environmental conditions. For example, studies have suggested a possible link between low sensitivity to drought and genotype [@ferrell1966; @teskey1987]. These studies show that although the interannual variation in density is under environmental control, its amplitude may be controlled by a genotype-environment interaction. Therefore, a question arises as to whether the amplitude of interannual variation is primarily under genetic control (heritability). Thus, in the context of climate change, it is crucial to know the level of heritability of the amplitude of interannual variation in wood density for the genetic selection of individuals with low amplitude of density variation. \

In this study, we evaluated the amplitude of the interannual variation of wood density in white spruce and its share of heritability in the interannual conditions of the environment. We therefore compare two growth sites containing genotyped trees in order to characterize the genetic (heritability) and environmental contribution to the interannual variation of wood density. Thus, we hope to select and reproduce white spruce trees families of interest that are both low-sensitivity to climates changes and genetically prone to make low amplitude of interannual variation of wood density. If the amplitude of variation in wood density is under genetic control, tree families with the smallest variation amplitudes will be identified for reforestation.

@@

The material used was sampled from an experimental provenance-progeny test plantation designed to assess the genetic merit of first generation selections, which had been subdivided into breeding groups and delineated by provenance or geographic region. The details of the setting up of the experiment are described in @beaulieu1996. Crosses were made according to a partial diallel mating design within each of the six different breeding groups or sublines [@Burdon1983], which were designed to limit future inbreeding to within group. Each parent tree has been used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups. Genetic testing was conducted in 1999 on 2-year-old nursery seedlings near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1). Trees were assigned to plots in rows of five trees / plot (spacing of 2 m × 2 m) [@beaulieu2014]. In this study, 93 full-brother families for a total of 2365 trees. The sampled trees were drawn from 18 blocks over the two sites and 9–14 individuals were sampled per full-sib cross. Trees used in this study are exclusively full-brother families.\

|Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |
|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|
|Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |
|Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |
Table: Location and climatic conditions* (over 1981–2010) of the two plantation sites.

@@

Wood cores were extracted à une épaisseur de 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground). Tree cores are extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself. These cores were used to determine ring width and wood density. Density was determined by scanning incremental cores from bark to bark along the radial face at a resolution of 25 $\mu$m in the QTRS-01X ring analyzer (Quintek Measurement Systems Inc., Knoxville, TN, USA).  A weighted mean density line was obtained by summing the product between the annual density measurements and the ratio of the respective radial areas to the total radial area. Wood cores were conditioned to about 7% moisture before scanning. Thus, the determination of the density of the wood makes it possible to calculate a quality index of it. From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines. Ring width determined through information of densitometry profiles.\

@@

## Model Development
A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations. Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018] (Eq.3). Previous studies have used this method to describe the density profile of different spruce species that have a similar profile[@franceschini2010; @gardiner2011]. 
However, when modeling density profiles we found that the variation between individual discs follows different patterns from the marrow to the bark between tree. We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others. To work around this problem, we chose to model each of the profiles separately. We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core . The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999].
As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores. A nonlinear equation was fit for each core to describe ring density profiles, producing different parameter estimates for each core. The radial pattern of wood density in white spruce can be described as follows:

@@

where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term. Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.  The graphic description in figure 2 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).

@@

## Determination of the wood density variability index
After subtraction of the typical density, it is observed that the residues remain variable. Knowing that the low frequency signal is subtracted, this variation can only be due to the high frequency signal which is the interannual variation. As explained earlier in the introduction, low interannual variability is an important criterion for wood quality. In order to identify tree families with less interannual variability in density, we have created a viability index.
The variability index is an index to determine the uniformity in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions. The residues of our model were used to obtain the wood density variability index, which was determined using an equation. We subtracted the density residues values of year $n-1$ from year $n$. We were interested more specifically in the difference between the average density residues of annual rings of two consecutive years, no matter whether the difference was positive or negative. The absolute value was used for each difference to get rid of the sign. The index was estimated from the following formula:

@@

Where V is the variability index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.

## Spearman’s rank correlation coefficient (Spearman’s rho) 
Spearman’s rank correlation coefficient is a correlation coefficient to describe the link between two variables by means of a monotonic function. We therefore used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites. We calculated this coefficient for the variability index and the different parameters of the model. The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites and the three parameters to estimate from the model. The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho. We used a 95 percent confidence interval.

@@

 $$h_2= 2 \frac{\sigma^2_f}{\sigma^2_T}$$

Where $h_2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_T$ is total variance.
ASReml3 package [@Butler2018] in the R statistical programming environment [@R2018] was used to estimate trait genetic parameters based on the residual maximum likelihood.

@@

|Site                     |Parameter |    Min|   Mean|    Max|
|:------------------------|:---------|------:|------:|------:|
| Asselin (Témiscouata)   |$b_0$     | 257.26| 302.78| 369.78|
|                         |$b_1$     |  -0.93|   0.97|   5.36|
|                         |$b_2$     | -81.83|  17.02|  61.34|
| Saint-Casimir (Portneuf)|$b_0$     | 264.88| 304.40| 353.90|
|                         |$b_1$     |  -1.34|   1.09|   7.04|
|                         |$b_2$     | -50.87|  14.87|  61.34|
Table: Summary table of parameters

![Pith-to-bark radial pattern of breast height density by ring. Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px}.\

![Pith-to-bark radial pattern of breast height density by ring. Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px}. \

@@

|Site                     |  Min|  Mean|    Max|
|:------------------------|----:|-----:|------:|
|Asselin (Témiscouata)    |  0.0| 417.5| 1155.8|
|Saint-Casimir (Portneuf) | 54.9| 339.8|  746.0|
|Combined                 |  0.0| 375.9| 1155.8|
Table: Summary table of variability index

@@

|Rank correlation  |  Rho| P-value|
|:-----------------|----:|-------:|
|Density residus   | 0.57|8.7^-09^|
|Variability index | 0.08|    0.43|
|$b_0$             | 0.70|2.2^-16^|
|$b_1$             | 0.80|2.2^-16^|
|$b_2$             | 0.60|2.2^-16^|
Table: Sperman’s Rang correlation

@@

|Trait            |Combined sites |Asselin |Saint-Casimir |
|:----------------|:--------------|:-------|:-------------|
|Wood density     |0.55           | 0.58   |0.70          |
|Width            |0.23           |0.32    |0.43          |
|$b_0$            |0.20           |0.20    |0.3           |
|$b_1$            |0.06           |0.10    |0.06          |
|$b_2$            |0.04           |0.10    |0.03          |
|Variability index |1.10^-6^       |6.10^-6^|3.10^-6^      |
Table: Abstract of the heritability of the density, the ring width, parameters $b_0$, $b_1$, $b_2$ and the Variability index.

>>>>

Thousands hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019]. In order to avoid deforestation, natural and artificial (planting or seeding) regeneration are used in Canada to compensate for harvests and forest losses. Artificial regeneration is used on more than half of the forests harvested in Canada [@RNC]. One of the reasons for artificial regeneration is for better control of planted trees to meet the demands and requirements of forest industry. Indeed, artificial regeneration, could ideally be made by using better trees than the former ones, i.e. trees that growth faster, that are more resistant to perturbation and that are of better structural quality. Thus, genetic selection is one of the tools for artificial regeneration and is becoming a crucial and increasingly necessary part of the development of canadian forest [@Bonfils2004]. When used in forestry, genetic selection allows to select trees with characters of interest to cross or to reproduce them in order to obtain trees with desired properties for an improved development of the forests or an improved quality of the wood for human use [@zobel1984; @DRF2000; @nanson2004]. \

During the first half of the 20^th^ century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012]. Hence, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties of commercial species such as white spruce. For example, improvement programs by genetic selection are in place for white spruce *(Picea glauca [Moench] Voss)* [@DRF2000; @Mullin2011] which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015] in North America. In Quebec, several generations of seed orchards have already been established in white spruce. Thus, the current objectives of genetic selection for this species are to select trees with a higher wood quality because white spruce has already reached a level considered interesting for traits such as rapid growth and volume of wood. \

@mitchell1961 defined the quality of the wood as follows: "Wood quality is the resultant of physical and chemical characteristics possessed by a tree or a part of a tree that enable it to meet the property requirements for different end products". We can consider that it is composed  of two components: one is biological and one economical. Biologically, resistance to mortality by vigor is important to ensure the survival of the tree and the value of the forest. Economically, the quality of the wood for end-use is important for industries as it impact the value of the end products. Several traits were considered to assess the different component of wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006]. More specifically the low interannual variability of wood density and rings width has an high importance for stem vigor and wood quality. First of all, it is important to understand that, the interannual variation of wood density comes from two sources: the low-frequency radial profile and the high-frequency radial profile [@Nicault2010]. The low-frequency radial profile is the radial variation of the density of pith-to-bark with a high density during the first years of tree growth  decreases to a certain level and begins to increase again. In white spruce, this radial profile is the radial type II profile described by @panshin1980. These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] because this functioning is internal to the tree and is transmitted to future generations. The high-frequency radial profile is the amplitude of the interannual variation of wood density i.e. the density difference between two consecutive growing years (fig 1).

![Schematic curves showing various signals in a dendrochronological series: low frequency age signal and high frequency climate signal. (modified from @Nicault2010)](graph_table/signal.png){width=200px; height=300px}

This high frequency signal is related to climatic conditions [@Nicault2010]. Therefore, significant inter-annual fluctuations in climate lead to an increase in inter-annual variation in wood density. In the coming years, the range of year-to-year variation in wood density may change due to climate changes. Indeed, wood density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a low variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008]. For example, @suarez2004 found that at the tree scale, individuals with variable growth were more prone to die of drought than more regularly growing trees. These studies have shown that the steady growth and therefore the low interannual variation wood density allows the tree to be more vigorous and therefore less prone to related mortality and environmental conditions. With regard to the economical context, interannual variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties. Indeed, low interannual variability of wood density affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002]. For example, the low interannual variation in wood density is appreciated for sawnwood because sawyers place a high value on the low interannual variation wood density [Hassegawa et al (pending approval); @X]. It is therefore important from both a biological and an end-use points of view to have a wood with a low interannual density variation within the stem. Therefore, we understand that there might be a strong gain at selecting trees that are prone to produce wood with a low interannual variability of wood density to get a resistant and highly valuable forest in the next decades. For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Beaulieu2009; @Mullin2011]. For white spruce wood quality, one of the characteristics that one wishes to select is the amplitude of the interannual variation of wood density, provided that it has sufficiently interesting heritability for a genetic selection. \

Heritability is the probability that a trait of an individual is hereditarily transmitted by genes. It is a statistical information that serves to evaluate the part of genome and that of environmental conditions in the construction of the phenotype of a given population. Moreover, several studies have been focused on the understanding the genetic sources of phenotypic variations of wood quality traits (eg. density) and their heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016]. Previous studies have shown that wood density is a trait under strong genetic control and therefore have a strong heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006]. As mentioned above, the low-frequency radial variation is age-related and under genetic control with strong heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016]. However, we did not find any studies on heritability of amplitude of interannual variation of wood density in white spruce.\

Studies have linked annual environmental conditions such as soil water availability, climatic conditions and interannual variability wood density (high frequency signal) [@becker1995; @huang2007; @Nicault2010; @zaehle2014]. Indeed, long-term growth analyzes have shown a link between wood density and local annual conditions such as summer temperature in white spruce [@darrigo1992] and other species [@becker1995; @huang2007; @zaehle2014]. However, the results of several studies suggest a genetic control of density sensitivity to environmental conditions. For example, studies have suggested a possible link between low sensitivity to drought and genotype [@ferrell1966; @teskey1987]. These studies show that although the interannual variation in density (high frequency signal) is under environmental control, its amplitude may be controlled by a genotype-environment interaction. Therefore, a question arises as to whether the amplitude of interannual variation is primarily under genetic control (heritability) or environmental. Knowing that the interannual variation in density is controlled by the annual environmental conditions, in the context of climatic changes that are likely to increase the interannual variation of environmental conditions, it is crucial to know whether the amplitude of variation in wood density is under genetic control. \

In this study, we evaluated the amplitude of the interannual variation (high frequency signal) of wood density in white spruce and its share of heritability in the interannual conditions of the environment. We therefore compare two growth sites containing genotyped trees in order to characterize the genetic (heritability) and environmental contribution to the amplitude of the interannual variation of wood density. Thus, we hope to select and reproduce white spruce trees families of interest that are both low-sensitivity to climates changes and genetically prone to make low amplitude of interannual variation of wood density. If the range of variation in wood density is under genetic control, tree families with the smallest variation ranges can be identified for reforestation.

@@

The material used was sampled from an experimental provenance-progeny test plantations designed to assess the genetic merit of first generation selections, which had been subdivided into breeding groups and delineated by provenance or geographic region. The details of the setting up of the experiment are described in @beaulieu1996. Crosses were made according to a partial diallel mating design within each of the six different breeding groups or sublines [@Burdon1983], which were designed to limit future inbreeding to within group. Each parent tree has been used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups. Genetic testing were set up in 1999 on 2-year-old seedlings from these experimental plantations near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1). Trees were assigned to plots in rows of five trees / plot (spacing of 2 m × 2 m) [@beaulieu2014]. In this study, we were used 93 full-brother families for a total of 2365 white spruce trees. The sampled trees were drawn from 18 blocks over the two sites and 9–14 individuals were sampled per full-sib cross. Trees used in this study are exclusively full-brother families.\

|Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |Ecological region      |
|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|:----------------------|
|Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |balsam fir-yellow birch|
|Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |maple–basswood         |
Table: Location and climatic conditions* (over 1981–2010) of the two plantation sites.

@@

Wood cores were extracted to a thickness of 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground). Tree cores are extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself. These cores were used to determine ring width and wood density. Density was determined by scanning incremental cores from pith-to-bark along the radial face at a resolution of 25 $\mu$m in the QTRS-01X ring analyzer (Quintek Measurement Systems Inc., Knoxville, TN, USA).  A weighted mean density line was obtained by summing the product between the annual density measurements and the ratio of the respective radial areas to the total radial area. Wood cores were conditioned to about 7% moisture before scanning. Thus, the determination of the density of the wood makes it possible to calculate a quality index of it. From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines. Ring width determined through information of densitometry profiles. The transition between earlywood and the latewood is determined using the "Unix" script. This script used a transition density threshold ($D_t$) determined for each ring as follows:

$$D_t= \frac{(D_{min}+D_{max})}{2}$$
Where $D_t$ is ring density threshold, $D_{min}$ is ring minimum density and $D_{max}$ is ring maximum density. Latewood started when these conditions were met:

1. $D_x < D_t$

2. $D_{x+1} >= D_t$

3. $D_{x+2} >= D_t$

Where $D_x$ is current reading and $D_t$ is next reading. \

@@

## Model Development
A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations. Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018] (Eq.3). Previous studies have used this method to describe the density profile of different spruce species that have a similar profile [@franceschini2010; @gardiner2011]. 
However, when modeling density profiles we found that the variation between individual discs follows different patterns from the pith-to-bark between tree. We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others. To work around this problem, we chose to model each of the profiles separately. We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core . The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values in addition to the explanatory variables, fit indices (R^2^) [@Pape1999].
As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores. A nonlinear equation was fit for each core to describe ring density profiles, producing different parameter estimates for each core. The radial pattern of wood density in white spruce can be described as follows:

@@

where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term. Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0. The graph in figure 2 describes the three parameters of our model and allows us to understand how each parameter influences our model (Fig 2).

@@

## Determination of interannual variation amplitude of wood density index
After subtraction of the typical density, it is observed that the residues remain variable. Knowing that the low frequency signal is subtracted, this variation can only be due to the high frequency signal which is the interannual variation. As explained earlier in the introduction, low interannual variation is an important criterion for wood quality. In order to identify tree families with low interannual variation amplitude in density, we have created a viability index.
The interannual variation amplitude of wood density index is an index to determine the variability in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions. The residues of our model were used to obtain the interannual variation amplitude of wood density index, which was determined using an equation. We subtracted the density residues values of year $n-1$ from year $n$. We were interested more specifically in the difference between the average density residues of annual rings of two consecutive years, no matter whether the difference was positive or negative. The absolute value was used for each difference to get rid of the sign. The index was estimated from the following formula:

@@

Where V is the interannual variation amplitude of wood density index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.

## Spearman’s rank correlation coefficient (Spearman’s rho) 
Spearman’s rank correlation coefficient is a correlation coefficient to describe the link between two variables by means of a monotonic function. We therefore used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites. We calculated this coefficient for the interannual variation amplitude of wood density index and the different parameters of the model. The elements for which we calculated Spearman’s rho are: the interannual variation amplitude of wood density index for each of the tree families between two study sites and the three parameters to estimate from the model. The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho. We used a 95 percent confidence interval.

@@

 $$h_2= 2 \frac{\sigma^2_f}{\sigma^2_t}$$

Where $h_2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_t$ is total variance.
ASReml3 package [@Butler2018] in the R statistical programming environment [@R2018] was used to estimate trait genetic parameters based on the residual maximum likelihood.

@@

|Site                     |Parameter |Min    |Mean   |Max    |
|:------------------------|:---------|------:|------:|------:|
| Asselin (Témiscouata)   |$b_0$     | 257.26| 302.78| 369.78|
|                         |$b_1$     |  -0.93|   0.97|   5.36|
|                         |$b_2$     | -81.83|  17.02|  61.34|
| Saint-Casimir (Portneuf)|$b_0$     | 264.88| 304.40| 353.90|
|                         |$b_1$     |  -1.34|   1.09|   7.04|
|                         |$b_2$     | -50.87|  14.87|  61.34|
Table: Summary table of parameters

![Pith-to-bark radial pattern of breast height density by ring. Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px} \

![Pith-to-bark radial pattern of breast height density by ring. Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px} \

@@

|Site                     |Min  |Mean  |Max    |
|:------------------------|----:|-----:|------:|
|Asselin (Témiscouata)    |  0.0| 417.5| 1155.8|
|Saint-Casimir (Portneuf) | 54.9| 339.8|  746.0|
|Combined                 |  0.0| 375.9| 1155.8|
Table: Summary table of variability index

@@

|Rank correlation  |Rho  |P-value |
|:-----------------|----:|-------:|
|Density residus   | 0.57|8.7^-09^|
|Variability index | 0.08|    0.43|
|$b_0$             | 0.70|2.2^-16^|
|$b_1$             | 0.80|2.2^-16^|
|$b_2$             | 0.60|2.2^-16^|
Table: Sperman’s Rang correlation

@@

|Trait             |Combined sites |Asselin |Saint-Casimir |
|:-----------------|:--------------|:-------|:-------------|
|Wood density      |0.55           |0.58    |0.70          |
|Width             |0.23           |0.32    |0.43          |
|$b_0$             |0.20           |0.20    |0.3           |
|$b_1$             |0.06           |0.10    |0.06          |
|$b_2$             |0.04           |0.10    |0.03          |
|Variability index |1.10^-6^       |6.10^-6^|3.10^-6^      |
Table: Abstract of the heritability of the density, the ring width, parameters $b_0$, $b_1$, $b_2$ and the Variability index.
